Bayesian Nonparametric Neural Image Segmentation

### 2013-02-06 latex/submission.tex
revised paper

\listoffixmes

\title{Bayesian Nonparametric Neural Image Segmentation}
\author{Kui Tang and Frank Wood}
\date{\today}
\maketitle
\begin{abstract}
We develop a Bayesian nonparametric method to segment neurons in electron micrograph images of brain tissue. Our model imposes spatial locality on arbitrary features, automatically determines the number of cells in an image, and can be used in both unsupervised and semi-supervised settings. We introduce a novel geodesic distance which adaptively adjust segmentation to respect detected cell boundaries. Using data collected by Kasthuri et al., we show that our model can obtain plausible segmentations with no training data, while receiving a significant boost in its presence. Finally, the Bayesian nonparametric model offers a natural extension to 3 dimensions, where neural columns of segments \emph{through} slices are modelled as coagulation-fragmentation processes, paving the road for a fully Bayesian solution to the connectomics problem.
\end{abstract}

@@

\section{Related Work}
Automated techniques generally begin by finding the cell boundaries---a highly distinctive feature. Because cells are usually disconnected, segmentation can be reduced to binary classification of whether voxels are inside or outside a cell. Segmentation can be done directlyin voxel space. Jain et al. \cite{Jain2007} achieved an 8.0\% test error with a 34,000 convolutional neural network, a statistically significant improvement other the 18.95\% error of thresholding, while MRFs and CRFs do no better than thresholding. Andres et al. obtain similar performance, at 8\% misclassification, using the watershed algorithm over a probability map of a random forest with hand-engineered features \cite{Andres2008}. However, the dataset used by both used a special intracellular stain which made segmentation easier. The generality of these solutions are thus unclear.

More common is to segment in pixel space and track the identity of segmented bodies in the vertical dimension. This mitigates the poor vertical resolution of EM images \cite{Kaynig2010, Briggman2006}. Jurrus et al. \cite{Jurrus2008} first obtain 2D segmentations with watershed after preprocessing linear diffusion filtering (NB: learn what this is!), and construct a weighted graph of these segments, with edges between adjacent sections with edge costs favoring high cross-correlation of segments low lateral displacement. Each neural process is thus a shortest path through this graph, computed by Dijkstra's algorithm. This cost function is robust to merging and loss of sections, but because Dijkstra's algorithm finds a single path, it cannot model branching. Kaynig et al. \cite{Kaynig2010a} presuppose segmentation and represent the image as a graph of regions, like Jurrus. Pairwise features are represented as edge weights, and a random forest classifier is trained to output a scalar similarity score for each pair. Then a probabilistic model encoding geometric constraints with cross-section correspondences as hidden variables is trained with EM. Finally, the maximum-likelihood result is the 3D reconstruction. These pairwise correspondences are fed to an agglomerative clustering algorithm, constructing the neural processes from ground-up in the same manner as a human. Unlike in Jurrus, processes may merge. However, each region is explicitly limited to two correspondences: one above and below, so processes still cannot branch.

Feature engineering significantly shapes the state of the art. Venkataraju et al. \cite{Venkataraju2009} generated 100 Hessian neighborhood features at each pixel and classified Jurrus's dataset with AdaBoost. Lucchi et al. \cite{Lucchi2010} develop ray descriptors, which capture shape features without a predefined shape model. These features are rotation, scale, and translation invariant, but strongly discriminate the types of objects observed in EM images. There features were used in SVMs to classify interior and boundary points of mitochondria, which are considerably harder to classify than other cellular structures \cite{Kaynig2010b}, and a graph cut found the final 2D segmentation with 98\% accuracy. Kaynig et al. encode perceptual grouping constraints---rules of thumb used by human experts to group neuron structures---as pairwise interaction terms in an energy function solved by graph cut. Minimizing the energy function gives a segmentation, and identifying maximum overlapping regions between slices is sufficient to obtain a 3D reconstruction

@@

\section{Model and Features}

\begin{figure}
\centering

@@

\label{sec:model}
We present our model in Figure~\ref{fig:model}. Below, we explain our feature representation and how we overcome the aforementioned difficulties. First, the number of cells must be inferred from the image itself. To do so, we employ a mixture Dirichlet process, discussed in section~\ref{sec:mdp}. Further, the visual complexity and noise of electron micrograph images do not yield the clear separations required for assumptions 2 and 3. We fix this with Radon-like features to separate cell interiors and boundaries in section~\ref{sec:radon}. Finally, our model encodes spatial locality in two novel ways: in the Gaussian covariance which bias class membership to Euclidean-close points, and in the geodesic distances which penalizes boundary crossings, discussed in section~\ref{sec:geodesic}.

@@

\begin{itemize}
\item $\+\pi \sim \text{DirichletProcess}(\alpha)$
\item $k | \+\pi \sim \text{Discrete}(\+\pi)$
\item If $k$ is a new (empty) cluster:
  \begin{itemize}
  \item $\Lambda_k | \Lambda_0, \nu \sim \text{Wishart}(\Lambda | \Lambda_0, \nu)$ (in the notation of \cite{Murphy2003})
  \item $\+\mu_k | \kappa , \Lambda_k \sim \mathcal{N}\left( \+\mu_k | \+\mu_0 , (\kappa\Lambda)^{-1} \right)$
  \item $\gamma_k | \beta_0, \gamma_0 \sim \text{Gamma}(\gamma_k | \beta_0, \text{rate} = \gamma_0)$
  \end{itemize}
\item $\+x | \+\mu_k , \Lambda_k, \gamma_k \propto \mathcal{N}(\+x_{1:4} | \+\mu_k , \Lambda_k) \text{Exponential}\left( d(\+x_{1:2}, \+\mu_{k,1:2}) | \text{rate} = \gamma_k \right) \label{eq:xlike}$ (in the notation of section~\ref{sec:geodesic})
\end{itemize}

\subsection{Mixture Dirichlet Processes}
\label{sec:mdp}
We begin with a finite mixture model shown in Figure~\ref{fig:finitemix} and derive the mixture Dirichlet process as a limiting case. This follows the derivation first given by \cite{Rasmussen2000}.

The indicators $z_n$ are drawn from the discrete distribution $\+\pi$, a $K$-dimensional vector which is in turn is drawn from a symmetric Dirichlet distribution parameterized by $\alpha$: $$p \left( \+\pi | \alpha, K \right) = \frac{ \Gamma ( \alpha ) }{ \Gamma ( \alpha/K ) ^K} \prod_{k=1}^K \+\pi_k^{\alpha/K - 1}.$$ Because the Dirichlet and discrete distributions are conjugate, we can marginalize $\+\pi$. Let $\+{z}_{-n}$ denote the vector of all indicators excluding element $n$ and $N_{k,-n}$ denote the number of points assigned to class $k$ excluding point $n$. A standard result \cite{Rasmussen2000} yields the posterior $$p(z_n = k | \+{z}_{-n})  = \frac{ N_{k,-n} + \alpha/K }{ N - 1 + \alpha }$$ which makes Gibbs sampling easy.

Now consider $K \rightarrow \infty$. First, the term $\alpha / K \rightarrow 0$. Given finitely many observations, we have only finitely many components $k$ with $N_{k,-n} > 0$. Therefore, the density 
\[
p(z_{n}=k|\+{z}_{-n},\alpha)=\begin{cases}
\frac{N_{k,-n}}{N-1+\alpha} & N_{k,-n}>0\\
\frac{\alpha}{N-1+\alpha} & \text{otherwise}
\end{cases} \label{eq:mdp}
\]
satisfies the limiting conditions and correctly normalizes to 1 when summed over $k = 1, \ldots, \infty$. We now say that the mixing distribution $\+\pi$, which is now infinite-dimensional, is drawn from a \emph{Dirichlet process}. Since conjugacy allows us to eliminate $\+\pi$ from computations, Gibbs sampling from an infinite mixture is just as efficient as for a finite mixture case.

@@

Ideally, for a mixture model, pixel intensities within a cell should be homogeneous. But the diversity of intracellular structures and significant imaging noise yield data far from this ideal. To correct this, we augment raw pixel intensities with $\emph{Radon-like feature}$ responses, which have proven useful for connectomics segmentation \cite{Kumar2010}. \fxwarning{Make this clearer... Even the original paper doesn't have a concise description}. A Radon-like feature aggregates a function of image intensity over regions defined by structural constraints, such as edges. Each feature is measured from a single direction, which we can represent as a \emph{scan line} $\+{l}(t) : [0, 1] \rightarrow \+{R}^2$. We represent the structure information as a set of points $\{ t_1, \ldots, t_m \}$ along line $\+{l}$. We follow the practice in \cite{Kumar2010} to evaluate Radon-like features along uniform angular intervals (we find 72 to avoid most discretization artifacts) and take the mean response from all directions. Of course, using the entire vector of feature responses with a multivariate likelihood is also possible. Moreover, in \cite{Kumar2010} as well as this paper, we will use the filter response $$R(x,y) = \max_{\sigma, \phi} \Delta G(\sigma, \phi) \ast I(x,y) $$ where $\Delta G(\sigma, \phi)$ is the Gaussian second derivative filter at scale $\sigma$ and orientation $\phi$. We use $R$ to (1) obtain knots, by passing to a Canny edge mapper, and (2) use as input to extraction functions (discussed below). Unlike other methods, we use the noisy edge response only as a guide for additional feature extraction, not as a testimony for the cell segments themselves.

Let $T(I, x, y)$ be an $\emph{extraction function}$ which maps an image and $xy$-coordinates to a scalar. Suppose a line $\+{l}$ is given and $(x, y) = \+{l}(t)$ for some $t \in [t_i, t_{i+1}]$. Then the Radon-like response for image $I$ at $(x, y)$ in direction $\+{l}$ is given by $$\Psi \left( I,\+{l},t \right) = T \left( I, \+{l}(t) \right).$$ We will use the extraction function proposed in \cite{Kumar2010} for cell background segmentation $$T_{\text{bg}}(I, \+{l}(t)) = \min_{t \in [t_i, t_{i+1}]} \left\{ I \left( \+{l}(t) \right) \right\}$$ which simply returns the dimmest pixel between any two knots. Because knots denote edges, this feature tends to homogenize pixel intensities within a cell.

Finally, we want pixels to not cross cell boundaries. The first step in encoding this constraint is to use a boundary-enhancing extraction function \cite{Kumar2010} $$T_{\text{bd}} = \frac{\int_t^{t+1} R\left( \+{l}(\tau) \right) d\tau }{\left\Vert \+{l}(t_{i+1}) - \+{l}(t_{i}) \right\Vert_2 }.$$ This function assigns to each pixel the average value of the Gaussian second-derivative response. Recall that the GSD tends to enhance boundaries, the most salient of which are captured by the Canny detector as knots. Thus, $T_{\text{bd}}$ smoothes the spurious GSD responses, avoiding oversegmentation. \fxwarning{Tighten up wording here.}Further, in connectomics data, cell membranes tend to form thick boundaries. But classic edge detectors work best at single-pixel gradients and return two edges for a thick boundary. In particular, the Canny detector produces two knots on either side of a cell boundary. In the GSD response $R$, cell boundaries are consistently bright, so the integral in $T_{\text{bd}}$ returns a distinctly brighter response on boundaries, regardless of thickness. We use this boundary enhancement information in the next section.

\subsection{Geodesic Distance}
\label{sec:geodesic}
\begin{figure}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_small}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_dist}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_thresh}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_hist}
\caption{Boundary Radon-like feature response; red marks the center from which the remaining pictures are measured. Grayscale map of geodesic distance (of superpixels). Threshold of geodesic distance $< 0.1$. Distribution of geodesic distances about the red center. Although the separation between interior and exterior is easy effort, BASELINE BLAH}\label{fig:geodesic}
\end{figure}
Figures~\ref{fig:radon} and~\ref{fig:geodesic} illustrates that cells need not be elliptical, or even convex. All that can be said is that they are contained in a closed loop that is generally gives a distinct Radon-like boundary response. Therefore, shape constraints based on Euclidean distances are not justified. Instead, we want to encode the a priori belief that "the cell body should not spill outside its membrane." Let us quantify this statement. Construct an undirected graph with pixels as vertices and pairs of adjacent pixels as edges ($(x,y)$ is an edge if $y$ is in the 1-pixel neighborhood of $x$). Each edge has a weight $w(i,j) = \left| b_i - b_j \right|$ where $b_k$ is the boundary Radon-like response at pixel $k$. We define the \emph{geodesic distance} $d(\+{x}_i, \+{x}_j)$ to be the total weight of the shortest path between pixels $i$ and $j$, which is computed with Johnson's algorithm. In practice, the $O(N^2)$ storage is prohibitive, so we instead we downsample the image into SLIC superpixels \cite{Achanta2012} and compute superpixel-based distances. Further, because individual cells are tiny relative to an entire slice, only local distances are needed, because far points must be outside the current cell. All paths crossing a cell boundary must traverse pixels with low, high, and again low boundary feature responses. Thus, the shortest path wil also be one with a high weight. So if pixels $i$ and $j$ do not belong to the same cell, defined by the region enclosed by a loop of pixels with high boundary feature response, then $d(\+{x}_i, \+{x}_j)$ is high.

@@

\section{Inference}
Inference is performed by Gibbs sampling, using Radford Neal's Algorithm 8 for non-conjugate priors \cite{Neal2000}. This is because the likelihood of $\+x$, which depends on the function $d$, is not a known conjugate-parametric form.\fxwarning{I'm not sure this is the right explanation.} We collapse $\+\pi$, but we do not collapse the other latent variables. Denote by $X$ the set of all $\+x_n$ and by $X_k = \left\{ \+x_n \in X : z_n = k \right\}$ and $N_k = | X_k |$. Then we derive the conditional posteriors by finding the Markov blankets of each node and using Bayes' rule. But because the likelihood of $\+x_n$ is note a standard parametric form, conditionals involving $X)k$ and $\+\mu_k$ cannot be sampled explicitly; we can only compute their likelihoods can be computed up to proportionality and resort to Metropolis-Hastings sampling.
\begin{align}\label{eq:conds}
p(z_n = k | \+\pi, \+x_n, \+\mu_k, \Lambda_k) &= p(z_n = k | \+\pi) \mathcal{N}(\+x_n | \+\mu_k, \Lambda_k) \label{eq:z}\\
p(\gamma_k | X_{z=k} , \beta_0, \gamma_0) &= \text{Gamma}\left( \gamma_k | \beta_0 + N_k, \gamma_0 + \sum_{\+x \in X_k} \+x \right) \label{eq:gamma}\\
p(\Lambda_k | X_k, \+\mu_k, \gamma_k, \Lambda_0, \nu) &\propto  p(X_k | \+\mu_k, \Lambda_k, \beta_k) \text{Wishart}(\Lambda_k | \Lambda_0, \nu) \label{eq:lambda}\\
p(\+\mu_k | X_k, \Lambda_k, \gamma_k, \+\mu_0, \kappa) &\propto p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \mathcal{N}(\+\mu_k | \+\mu_0, (\kappa\Lambda)^{-1}) \label{eq:mu}
\end{align}
Note that $\+\pi$ remains collapsed; to evaluate \eqref{eq:z}, see \eqref{eq:mdp}.

\begin{algorithm}
\caption{Gibbs sample iteration}\label{alg:gibbs}
\begin{algorithmic}
\Require The current state of the Markov chain, $\Theta = \left\{ N, K, Z, \left\{ N_k, \+\mu_k, \Lambda_k, \gamma_k \right\}_{k=1}^K  \right\}$
\Ensure The distribution of the next state $\Theta^{*}$ preserves the Markov chain's invariant distribution.
\Statex

\For{$n := 1:N$}
  \If{$N_{z_n} = 1$}
    \State $\left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right) := \left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right)$ 
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+2, \ldots, K+m\}$
  \Else
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+1, \ldots, K+m\}$
  \EndIf
  \State Draw $z_n$ from 
  \[
  p(z_{n}=k|\+{z}_{-n},\+x_n,\+\mu_k,\Lambda_k,\gamma_k,\alpha)=\begin{cases}
  \frac{N_{k,-n}}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \beta_k) & k < K\\
  \frac{\alpha / m}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \beta_k) & \text{otherwise}
  \end{cases} \label{eq:nealalg8}
  \]
  \State Delete parameters for classes $k$ with $N_k = 0$ and relabel the $z_n$ to remove gaps.
\EndFor

\For{$k := 1:K$}
  \State Draw $\gamma_k$ using \eqref{eq:gamma}
  \State Metropolis-Hastings sample $\Lambda_k$ using likelihood \eqref{eq:lambda} and proposal $\text{Wishart}(\Lambda^{*} | p \Lambda, \nu)$ for some $p > 1$.
  \State Metropolis-Hastings sample $\+\mu_k$ using likelihood \eqref{eq:mu} and proposal $\mathcal{N}( \+\mu^{*} | C )$.
\EndFor

@@

\subsection{Proposal Distributions}
Care is required to select proposal distributions for \eqref{eq:mu}, because the likelihood for $\+x_n$ is discontinuous in $\+\mu_k$, since it relies on the discrete function $d$ which arises from graph distances. The lack of reliable gradient information frustrates the smooth exploration of low-probability areas. However, discontinuities only occur at superpixel boundaries, and sharp discontinuities only at the transition between cell boundary and interior. Therefore, we use a diagonal-covariance Gaussian proposal whose $xy$ coordinate marginal forms a 95\% confidence ellipse that covers the neighboring superpixels. SLIC superpixels have roughly uniform size, even if shapes differ. Thus, a circle can be guaranteed to cover some neighbors of a superpixel. Thus, we propose with the following covariance
\[
C = \left(\begin{array}{cccc}
    \sigma^{2}\\
     & \sigma^{2}\\
      &  & \tau^{2}\\
       &  &  & \tau^{2}
     \end{array}\right)
\]
where $\sigma$ is the diameter of a SLIC superpixel and $\tau = 1/2$, because the third and fourth features are constrained to lie in $[0, 1]$.

@@

>>>>

\title{Bayesian Nonparametric Neural Image Segmentation}
\author{Kui Tang and Frank Wood}
\date{\today}
\maketitle
\begin{abstract}
We develop a Bayesian nonparametric method to segment neurons in electron micrograph images of brain tissue. Our model imposes spatial locality on arbitrary features, automatically determines the number of segments in an image, and can be used in both unsupervised and semi-supervised settings. We introduce a novel geodesic distance feature which adaptively adjust segmentation to respect detected cell boundaries. Using data collected by Kasthuri et al., we show that our model can obtain plausible segmentations with no training data, while receiving a significant boost in its presence. Finally, our Bayesian nonparametric naturally extends to 3 dimensions, where neural columns of segments \emph{through} slices are modelled.
\end{abstract}

@@

\section{Related Work}
Automated techniques generally try to find cell boundaries---a highly distinctive feature. Because cells are usually disconnected, segmentation can be reduced to binary classification of whether voxels are inside or outside a cell. Segmentation can be done directlyin voxel space. Jain et al. \cite{Jain2007} achieved an 8.0\% test error with respect to hand-labelled ground truth with a convolutional neural network. This beats the 18.95\% error of thresholding; MRFs and CRFs do no better. Andres et al. obtain similar performance over the same dataset using the watershed algorithm over a probability map of a random forest with hand-engineered features \cite{Andres2008}. However, their dataset used a special intracellular stain which made segmentation easier. The generality of these solutions are thus unclear.

More common is to segment in pixel space and track the identity of segmented bodies in the vertical dimension. This mitigates the poor vertical resolution of EM images \cite{Kaynig2010, Briggman2006}. Jurrus et al. \cite{Jurrus2008} first obtain 2D segmentations with watershed after preprocessing, and construct a graph of segments across layers. Edges connect all segments that overlap in the plane, and weights favor segments with high cross-correlation low lateral displacement. Each neural process is a shortest path through this graph, computed by Dijkstra's algorithm. This cost function is robust to merging and losing of sections. But because Dijkstra's algorithm finds a single path, it cannot model branching. Kaynig et al. \cite{Kaynig2010a} presuppose segmentation and also represent the image as a graph of regions. Edge weights store pairwise features, and a random forest is trained to output a scalar similarity score for each pair. Then a probabilistic model encoding geometric constraints with cross-section correspondences as hidden variables is trained with EM. The maximum-likelihood result is the 3D reconstruction. These pairwise correspondences are fed to an agglomerative clustering algorithm, constructing the neural processes from ground-up in the same manner as a human. Unlike in Jurrus, processes may merge. However, each region is explicitly limited to two correspondences: one above and below, so processes still cannot branch.

Feature engineering significantly shapes the state of the art. Venkataraju et al. \cite{Venkataraju2009} generated 100 Hessian neighborhood features at each pixel and classified Jurrus's dataset with AdaBoost. Lucchi et al. \cite{Lucchi2010} develop ray descriptors, which capture shape features without a predefined shape model. These features are rotation, scale, and translation invariant, but strongly discriminate the types of objects observed in EM images. There features were used in SVMs to classify interior and boundary points of mitochondria, considerably harder to classify than other cellular structures \cite{Kaynig2010b}, and a graph cut found the final 2D segmentation with 98\% accuracy.

@@

\section{Model and Features}
\label{sec:model}
\begin{figure}
\centering

@@

Although the confidence regions of Gaussians are ellipsoids in feature space, it is important to recall that decision boundaries need not be ellipsoidal at all. This is particularly true when the data are inseparable, i.e.~when each point has a high probability of belonging to several nearby clusters. This is the case in our neural image data: there are many cell segments, and a pixel's membership has significant uncertainty. The actual decision surface is determined by the relative likelihoods of the point for its various neighboring clusters and can be quite complicated.

We present our model in Figure~\ref{fig:model}. Below, we explain our feature representation and how we overcome the aforementioned difficulties. First, the number of cells must be inferred from the image itself. To do so, we employ a Dirichlet process mixture, discussed in section~\ref{sec:mdp}. Further, the visual complexity and noise of electron micrograph images do not yield the clear separations required for assumptions 2 and 3. We fix this with Radon-like features \cite{Kumar2010} to separate cell interiors and boundaries in section~\ref{sec:radon}. Finally, our model encodes spatial locality in two novel ways: in the Gaussian covariance which bias class membership to Euclidean-close points, and in the geodesic distances which penalizes boundary crossings, discussed in section~\ref{sec:geodesic}.

@@

\begin{itemize}
\item $\+\pi \sim \text{DP}(\alpha)$
\item $z | \+\pi \sim \text{Discrete}(\+\pi)$
\item For $k = 1, 2, \ldots$
\begin{itemize}
\item $\Lambda_k | \Lambda_0, \nu \sim \text{Wishart}(\Lambda | \Lambda_0, \nu)$
\item $\+\mu_k | \kappa , \Lambda_k \sim \mathcal{N}\left( \+\mu_k | \+\mu_0 , (\kappa\Lambda)^{-1} \right)$
\item $\gamma_k | \beta_0, \gamma_0 \sim \text{Gamma}(\gamma_k | \beta_0, \gamma_0)$
\end{itemize}
\item $\+x | z = k, \left\{ \+\mu_k , \Lambda_k, \gamma_k \right\}_{k=1}^{\infty} \propto \mathcal{N}(\+x_{1:4} | \+\mu_k , \Lambda_k) \text{Exponential}\left( d(\+x_{1:2}, \+\mu_{k,1:2}) | \gamma_k \right) \label{eq:xlike}$
\end{itemize}

We use the rate parameterization of the Gamma distribution: $\text{Gamma}(x | \beta, \gamma) = \frac{\gamma^\beta}{\Gamma(\beta)} x^{\beta-1} \exp{(-\gamma x)}$. The exponential distribution is just the Gamma distribution with $\alpha = 1$. The degrees of freedom of the Wishart distribution is $\nu$. The subscript $1:d$ denotes extracting the first $d$ dimensions of the vector.

@@

Ideally, for a mixture model, pixel intensities within a cell should be homogeneous. But the diversity of intracellular structures and significant imaging noise yield data far from this ideal. To correct this, we augment raw pixel intensities with $\emph{Radon-like feature}$ responses, which have proven useful for connectomics segmentation \cite{Kumar2010}. A Radon-like feature aggregates a function of image intensity over regions defined by structural constraints. Given emph{scan line} $\+{l}(t) : [0, 1] \rightarrow \+{R}^2$ denoting a direction along the image, we integrate an \emph{extraction function} piecewise, with segments defined by \emph{knots}, which are structurally salient points of the image. In particular, we determine knots as the poitns classified as edges by using a Canny edge mapper on the filter response $$R(x,y) = \max_{\sigma, \phi} \Delta G(\sigma, \phi) \ast I(x,y) $$ given in \cite{Kumar2010}, where $\Delta G(\sigma, \phi)$ is the Gaussian second derivative filter at scale $\sigma$ and orientation $\phi$. This edge response is noisy, but unlike other methods, we use it the noisy edge response only as a guide for additional feature extraction, not as a testimony for the cell segments themselves.

For a given scan line, we write its knots as parametric coordinates along the line: $K_{\+l} = \{ t_1, \ldots, t_m \}$. We follow the practice in \cite{Kumar2010} to evaluate Radon-like features along uniform angular intervals (we find 72 sufficient to avoid most discretization artifacts) and take the mean response from all directions. Of course, using the entire vector of feature responses with a multivariate likelihood is also possible. Let $T(I, x, y)$ be an extraction function which maps an image and $xy$-coordinates to a scalar. Suppose a line $\+{l}$ is given and $(x, y) = \+{l}(t)$ for some $t \in [t_i, t_{i+1}]$. Then the Radon-like response for image $I$ at $(x, y)$ in direction $\+{l}$ is given by $$\Psi \left( I,\+{l},t \right) = T \left( I, \+{l}(t) \right).$$ We will use the extraction function proposed in \cite{Kumar2010} for cell background segmentation $$T_{\text{bg}}(I, \+{l}(t)) = \min_{t \in [t_i, t_{i+1}]} \left\{ I \left( \+{l}(t) \right) \right\}$$ which simply returns the dimmest pixel between any two knots. Because knots denote edges, this feature tends to homogenize pixel intensities within a cell.

Finally, we want pixels to not cross cell boundaries. The first step in encoding this constraint is to use a boundary-enhancing extraction function \cite{Kumar2010} $$T_{\text{bd}} = \frac{\int_t^{t+1} R\left( \+{l}(\tau) \right) d\tau }{\left\Vert \+{l}(t_{i+1}) - \+{l}(t_{i}) \right\Vert_2 }.$$ This function assigns to each pixel the average value of the Gaussian second-derivative response. Recall that the GSD tends to enhance boundaries, the most salient of which are captured by the Canny detector as knots. Thus, $T_{\text{bd}}$ smoothes the spurious GSD responses, avoiding oversegmentation. Further, in connectomics data, cell membranes tend to form thick boundaries. But classic edge detectors work best at single-pixel gradients and return two edges for a thick boundary. In particular, the Canny detector produces two knots on either side of a cell boundary. In the GSD response $R$, cell boundaries are consistently bright, so the integral in $T_{\text{bd}}$ returns a distinctly brighter response on boundaries, regardless of thickness. We use this boundary enhancement information in the next section.

\subsection{Geodesic Distance}
\label{sec:geodesic}
\begin{figure}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_small}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_dist}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_thresh}
\includegraphics[width=0.24\linewidth, clip]{JAN_bd_hist}
\caption{(1) Boundary Radon-like feature response; red marks the center from which the remaining pictures are measured. (2) Grayscale map of geodesic distance (of superpixels). (3) Threshold of geodesic distance $< 0.1$. Distribution of geodesic distances about the red center.\label{fig:geodesic}}
\end{figure}
Figures~\ref{fig:radon} and~\ref{fig:geodesic} illustrate that cells need not be elliptical, or even convex. All that can be said is that they are contained in a closed loop that is generally gives a distinct Radon-like boundary response. Therefore, shape constraints based on Euclidean distances are not justified. Instead, we want to encode the a priori belief that ``the cell body should not spill outside its membrane.'' Let us quantify this statement. Construct an undirected graph with pixels as vertices and pairs of adjacent pixels as edges ($(x,y)$ is an edge if $y$ is in the 1-pixel neighborhood of $x$). Each edge has a weight $w(i,j) = \left| b_i - b_j \right|$ where $b_k$ is the boundary Radon-like response at pixel $k$. We define the \emph{geodesic distance} $d(\+{x}_i, \+{x}_j)$ to be the total weight of the shortest path between pixels $i$ and $j$, which is computed with Johnson's algorithm. In practice, the $O(N^2)$ storage is prohibitive, so we instead we downsample the image into SLIC superpixels \cite{Achanta2012} and compute superpixel-based distances. Further, because individual cells are tiny relative to an entire slice, only local distances are needed, because far points must be outside the current cell. All paths crossing a cell boundary must traverse pixels with low, high, and again low boundary feature responses. Thus, the shortest path will also be one with a high weight. So if pixels $i$ and $j$ do not belong to the same cell, defined by the region enclosed by a loop of pixels with high boundary feature response, then $d(\+{x}_i, \+{x}_j)$ is high.

@@

\section{Inference}
Inference is performed by Gibbs sampling, using Radford Neal's Algorithm 8 for non-conjugate priors \cite{Neal2000}. This is because the likelihood of $\+x$, which depends on the function $d$, is not a known conjugate-parametric form.\fxwarning{I'm not sure this is the right explanation.} We collapse $\+\pi$, but we do not collapse the other latent variables. Denote by $X_{-n}$ the set of all $\+x_n$ with point $n$ removed, by $X_k = \left\{ \+x_n \in X : z_n = k \right\}$ and $N_k = | X_k |$. Then we derive the conditionals by finding the Markov blankets of each node and using Bayes' rule. But because the likelihood of $\+x_n$ is not a standard parametric form, conditionals involving $X_k$ and $\+\mu_k$ cannot be sampled explicitly; we can only compute their likelihoods up to proportionality and resort to Metropolis-Hastings sampling.
\begin{align}\label{eq:conds}
p(z_n = k | \+\pi, \+x_n, \+\mu_k, \Lambda_k) &\propto p(z_n = k | \+\pi) \mathcal{N}(\+x_n | \+\mu_k, \Lambda_k) \label{eq:z}\\
p(\gamma_k | X_k, \+\mu_k, \beta_0, \gamma_0) &= \text{Gamma}\left( \gamma_k | \beta_0 + N_k, \gamma_0 + \sum_{\+x \in X_k} d(\+x, \+\mu_k) \right) \label{eq:gamma}\\
p(\Lambda_k | X_k, \+\mu_k, \gamma_k, \Lambda_0, \nu) &\propto  p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \text{Wishart}(\Lambda_k | \Lambda_0, \nu) \label{eq:lambda}\\
p(\+\mu_k | X_k, \Lambda_k, \gamma_k, \+\mu_0, \kappa) &\propto p(X_k | \+\mu_k, \Lambda_k, \gamma_k) \mathcal{N}(\+\mu_k | \+\mu_0, (\kappa\Lambda)^{-1}) \label{eq:mu}
\end{align}
Note that $\+\pi$ remains collapsed; to evaluate \eqref{eq:z}, see \eqref{eq:mdp}.

\begin{algorithm}
\caption{Gibbs sample sweep}\label{alg:gibbs}
\begin{algorithmic}
\Require The current state of the Markov chain, $\Theta = \left\{ N, K, Z, \left\{ N_k, \+\mu_k, \Lambda_k, \gamma_k \right\}_{k=1}^K  \right\}$
\Statex

\For{$n := 1:N$}
  \If{$N_{z_n} = 1$}
    \State $\left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right) := \left( \+\mu_{K+1}, \Lambda_{K+1}, \gamma_{K+1} \right)$ 
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+2, \ldots, K+m\}$
  \Else
    \State Draw $\+\mu_j, \Lambda_j, \gamma_j$ from their priors (see section~\ref{sec:model}), for $j \in \{K+1, \ldots, K+m\}$
  \EndIf
  \State Draw $z_n$ from 
  \[
  p(z_{n}=k|\+{z}_{-n},X_{-n},\alpha,\left\{ \+\mu_k,\Lambda_k,\gamma_k\ \right\}_{k=1}^{\infty}) \propto \begin{cases}
  \frac{N_{k,-n}}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & k < K\\
  \frac{\alpha / m}{N-1+\alpha} p(\+x_n | \+\mu_k, \Lambda_k, \gamma_k) & \text{otherwise}
  \end{cases} \label{eq:nealalg8}
  \]
  \State Delete parameters for classes $k$ with $N_k = 0$ and relabel the $z_n$ to remove gaps.
\EndFor

\For{$k := 1:K$}
  \State Draw $\gamma_k$ using \eqref{eq:gamma}
  \State Metropolis-Hastings sample $\Lambda_k$ using likelihood \eqref{eq:lambda} and proposal $\text{Wishart}(\Lambda^{*} | b \Lambda, \nu)$ for some $b > 1$.
  \State Metropolis-Hastings sample $\+\mu_k$ using likelihood \eqref{eq:mu} and proposal $\mathcal{N}( \+\mu^{*} | C )$.
\EndFor

@@

\subsection{Proposal Distributions}
Care is required to select proposal distributions for \eqref{eq:mu}, because the likelihood for $\+x_n$ is discontinuous in $\+\mu_k$, since it relies on the discrete function $d$ which arises from graph distances. The lack of reliable gradient information frustrates the smooth exploration of low-probability areas. However, discontinuities only occur at superpixel boundaries, and sharp discontinuities only at the transition between cell boundary and interior. Therefore, we use a diagonal-covariance Gaussian proposal whose $xy$ coordinate marginal forms a 95\% confidence ellipse that covers the neighboring superpixels. SLIC superpixels have roughly uniform size, even if shapes differ. Thus, a circle can be guaranteed to cover some neighbors of a superpixel. Thus, we propose with the following covariance
\[
C = \left(\begin{array}{cccc}
    \sigma^{2} & & & \text{\huge 0}\\
     & \sigma^{2} & & \\
      &  & \tau^{2} &\\
    \text{\huge 0} &  &  & \tau^{2}
     \end{array}\right)
\]
where $\sigma$ is the diameter of a SLIC superpixel and $\tau = 1/2$, because the third and fourth features are constrained to lie in $[0, 1]$.

@@

\appendix
\section{Dirichlet Process Mixtures}
\label{sec:mdp}
We begin with a finite Gaussian mixture model and derive the Dirichlet process mixture as a limiting case. This follows the derivation first given by \cite{Rasmussen2000}.
